Continuous developmental changes in word recognition support language learning across early childhood

### 2025-03-13 paper/paper.Rmd
fix a citation rendering issue

By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours. In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change. Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [bergelson2023;@frank2017;@frank2021;@manybabies2020].

>>>>

By aggregating data from many pre-existing studies, we were able to overcome the limitations of prior investigations, which typically had sample sizes at least an order of magnitude smaller than ours. In contrast to individual studies, which typically have at best the statistical power to test one or two specific contrasts, our "big data" approach provided the sample sizes necessary to fit complex structural equation models and to compare different functional forms for developmental change. Because early language is so variable [@frank2021], these kinds of samples -- with thousands, rather than dozens of children -- are likely to be required to gain further insight into the psychometrics of early language learning [@bergelson2023;@frank2017;@frank2021;@manybabies2020].

### 2025-03-13 paper/paper.tex
fix a citation rendering issue

% Please add here a significance statement to explain the relevance of your work
\significancestatement{The efficiency with which children recognize
words provides a measurement of their real-time language processing,
which has been argued to be a key part of the language learning process.
Her, we use a large dataset of eye-tracking data from many different
experiments to map out the development of word recognition, finding that
this process becomes faster, more accurate, and less variable in the
period between one and six years of age, and that children who are
better at word recognition tend to have larger and faster growing
vocabularies. These data suggest that understanding language is a skill
that improves gradually with practice.}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  one |  two |  optional |  optional |  optional  }

@@

By aggregating data from many pre-existing studies, we were able to
overcome the limitations of prior investigations, which typically had
sample sizes at least an order of magnitude smaller than ours. In
contrast to individual studies, which typically have at best the
statistical power to test one or two specific contrasts, our ``big
data'' approach provided the sample sizes necessary to fit complex
structural equation models and to compare different functional forms for
developmental change. Because early language is so variable (2), these
kinds of samples -- with thousands, rather than dozens of children --
are likely to be required to gain further insight into the psychometrics
of early language learning
{[}\href{mailto:bergelson2023;@frank2017}{\nolinkurl{bergelson2023;@frank2017}};(2);(43){]}.

At the same time, our approach is both observational and exploratory.
Thus, we cannot untangle the range of different causal models that
explain the variation we observed. First, early word recognition skill
could lead to faster word learning, due to skilled children gleaning
more information from the same signal (22). But second, faster children
could also be faster due to their larger vocabulary and stronger lexical
representations. These two causal directions could also interact
reciprocally, leading to a ``rich get richer'' process in which children
with larger vocabularies process faster, and their faster processing
helps them increase their vocabulary size more rapidly. Finally, a third
shared factor -- perhaps general cognitive ability -- could underpin
both processes. Our cross-sectional data cannot distinguish these
hypotheses even in principle (44), and our longitudinal data are likely
too sparse to distinguish such complex causal models. Thus, these
questions remain an open target for dense longitudinal data collection.

@@

\bibitem[\citeproctext]{ref-manybabies2020}
\CSLLeftMargin{43. }%
\CSLRightInline{The ManyBabies Consortium (2020) {Quantifying sources of
variability in infancy research using the infant-directed speech
preference}. \emph{Advances in Methods and Practices in Psychological
Science} 3(1):24--52.}

\bibitem[\citeproctext]{ref-vanderweele2023dimensional}
\CSLLeftMargin{44. }%
\CSLRightInline{VanderWeele TJ, Batty CJ (2023) On the dimensional
indeterminacy of one-wave factor analysis under causal effects.
\emph{Journal of Causal Inference} 11(1):20220074.}

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{The efficiency with which children recognize
words provides a measurement of their real-time language processing,
which has been argued to be a key part of the language learning process.
Here, we use a large dataset of eye-tracking data from many different
experiments to map out the development of word recognition, finding that
this process becomes faster, more accurate, and less variable in the
period between one and six years of age, and that children who are
better at word recognition tend to have larger and faster growing
vocabularies. These data suggest that understanding language is a skill
that improves gradually with practice.}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  language development |  word recognition |  developmental
change |  skill acquisition |  eye-tracking  }

@@

By aggregating data from many pre-existing studies, we were able to
overcome the limitations of prior investigations, which typically had
sample sizes at least an order of magnitude smaller than ours. In
contrast to individual studies, which typically have at best the
statistical power to test one or two specific contrasts, our ``big
data'' approach provided the sample sizes necessary to fit complex
structural equation models and to compare different functional forms for
developmental change. Because early language is so variable (2), these
kinds of samples -- with thousands, rather than dozens of children --
are likely to be required to gain further insight into the psychometrics
of early language learning (2, 43--45).

At the same time, our approach is both observational and exploratory.
Thus, we cannot untangle the range of different causal models that
explain the variation we observed. First, early word recognition skill
could lead to faster word learning, due to skilled children gleaning
more information from the same signal (22). But second, faster children
could also be faster due to their larger vocabulary and stronger lexical
representations. These two causal directions could also interact
reciprocally, leading to a ``rich get richer'' process in which children
with larger vocabularies process faster, and their faster processing
helps them increase their vocabulary size more rapidly. Finally, a third
shared factor -- perhaps general cognitive ability -- could underpin
both processes. Our cross-sectional data cannot distinguish these
hypotheses even in principle (46), and our longitudinal data are likely
too sparse to distinguish such complex causal models. Thus, these
questions remain an open target for dense longitudinal data collection.

@@

\bibitem[\citeproctext]{ref-bergelson2023}
\CSLLeftMargin{43. }%
\CSLRightInline{Bergelson E, et al. (2023) Everyday language input and
production in 1,001 children from six continents. \emph{Proceedings of
the National Academy of Sciences of the United States of America}
120(52):e2300671120.}

\bibitem[\citeproctext]{ref-frank2017}
\CSLLeftMargin{44. }%
\CSLRightInline{Frank MC, Braginsky M, Yurovsky D, Marchman VA (2017)
{Wordbank: An open repository for developmental vocabulary data}.
\emph{Journal of Child Language} 44(3):677--694.}

\bibitem[\citeproctext]{ref-manybabies2020}
\CSLLeftMargin{45. }%
\CSLRightInline{The ManyBabies Consortium (2020) {Quantifying sources of
variability in infancy research using the infant-directed speech
preference}. \emph{Advances in Methods and Practices in Psychological
Science} 3(1):24--52.}

\bibitem[\citeproctext]{ref-vanderweele2023dimensional}
\CSLLeftMargin{46. }%
\CSLRightInline{VanderWeele TJ, Batty CJ (2023) On the dimensional
indeterminacy of one-wave factor analysis under causal effects.
\emph{Journal of Causal Inference} 11(1):20220074.}
